---
title: "Running ChromatogramsVis inside Galaxy"
author: "Gabriele Tomè, Johannes Rainer"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Running ChromatogramsVis inside Galaxy}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{ChromatogramsVis}
    %\VignetteDepends{BiocStyle,knitr}
---

```{r setup, echo = FALSE, results = 'asis', message=FALSE, include = FALSE}
BiocStyle::markdown()
knitr::opts_chunk$set(
  eval = FALSE   # all commands below are meant to be run in a terminal, not by knitr
)
```

# Introduction

`ChromatogramsVis` ships with a [Shiny](https://shiny.posit.co/) application for
the interactive visualisation of
[Chromatograms](https://bioconductor.org/packages/Chromatograms) object.
Besides running the app locally from R, the package can also be executed inside
**Galaxy Interactive Tool**.

[Galaxy](https://galaxyproject.org/) is a web-based platform for reproducible
data analysis. An *Interactive Tool* (IT) is a Galaxy tool that, instead of
producing output files and finishing, starts a long-running web application
(here, the Shiny app) inside a Docker container and exposes it to the user
through the Galaxy interface.

This vignette explains, step by step, how to:

1. build the Docker image that contains the app;
2. register the tool in a local Galaxy instance;
3. start Galaxy and launch the app;
4. troubleshoot the most common problems.

It is intended for people who want to **test or develop** the tool on their own
machine.

# How the pieces fit together

The package contains three elements that matter for Galaxy:

| Element | Location (package root) | Purpose |
|---|---|---|
| `Dockerfile` | `Dockerfile` | Recipe that installs R, the package and its dependencies, and starts the Shiny app. |
| Tool definition | `dev/interactivetool_chromatogramsvis.xml` | Galaxy XML file describing the tool: name, inputs, Docker image to use, port to expose. |
| `DESCRIPTION` | `DESCRIPTION` | Contains the package `Version`. |

The Docker image is tagged as

```
chromatogramsvis:<CONTAINER_VERSION>_<TOOL_VERSION>
```

where

- `TOOL_VERSION` is the `ChromatogramsVis` version;
- `CONTAINER_VERSION` is the value of the `@CONTAINER_VERSION@` token declared
  in the tool XML file.

The same tag must be referenced in the XML file (inside the `<container>`
element), otherwise Galaxy will not find the image you built. The commands
below compute the tag automatically, so the two always stay consistent as long
as you rebuild the image whenever either version changes.

# Requirements

Before starting, make sure that your machine has:

* **Docker**, installed and running, and your user allowed to run `docker`
  without `sudo` (check with `docker run hello-world`).
* **Python 3** (needed by Galaxy; `./run.sh` creates its own virtual
  environment, so no manual Python setup is normally required).
* The clone of `ChromatogramsVis` GitHub repository.


# Step-by-step instructions

## 1. Download the ChromatogramsVis source code (only first time)

Clone the package repository from GitHub:
```{bash}
git clone https://github.com/gabrieletome/ChromatogramsVis.git
```

> **Note.** This step has to be done only once. If the `ChromatogramsVis`
> directory already exists, skip it.

## 2. Download the Galaxy source code (only first time)

Galaxy is cloned *next to* the package directory. The clone is large and can
take several time.

```{bash}
## Download Galaxy source code in "./galaxy"
## (i.e. "../galaxy" when seen from ChromatogramsVis)
git clone https://github.com/galaxyproject/galaxy.git
```

> **Note.** This step has to be done only once. If the `galaxy` directory
> already exists, skip it.

## 3. Directory layout

At this point, the commands assume that `ChromatogramsVis` and `galaxy` are
**sibling directories**:

```
workspace/
├── ChromatogramsVis/
└── galaxy/
```

All the commands are executed in a terminal. Start from the root of the
`ChromatogramsVis` package:

```{bash}
cd /path/to/workspace/ChromatogramsVis
```

## 4. Build the Docker image

Return to the package directory and build the image.
What these lines do:

- `TOOL_VERSION=...` reads the package version from `DESCRIPTION`.
- `CONTAINER_VERSION=...` reads the container version from the
  `@CONTAINER_VERSION@` token in the tool XML file.
- `docker build ...` builds the image from the `Dockerfile` in the current
  directory (`.`) and tags it `chromatogramsvis:<CONTAINER_VERSION>_<TOOL_VERSION>`.

```{bash}
cd ChromatogramsVis

TOOL_VERSION=`grep Version DESCRIPTION | cut -d: -f 2 | tr -d " "`
CONTAINER_VERSION=`grep '<token name="@CONTAINER_VERSION@">' dev/interactivetool_chromatogramsvis.xml | cut -d'>' -f2 | cut -d'<' -f1`

docker build . -t chromatogramsvis:${CONTAINER_VERSION}_${TOOL_VERSION} -f Dockerfile
```

You can check the variables and the result with:

```{bash}
echo "${CONTAINER_VERSION}_${TOOL_VERSION}"
docker images chromatogramsvis
```

The first build can take several minutes because R and all package
dependencies must be installed. Subsequent builds are faster thanks to
Docker's layer cache.

> **Important.** Re-run this step every time you change the R code, the
> `Dockerfile`, the package `Version` or the container version token.
> Galaxy uses the image *as it is* in your local Docker; it does not rebuild it.

## 5. Copy the tool definition into Galaxy

Galaxy keeps its interactive tools in `tools/interactive/`. The file name
(`interactivetool_chromatogramsvis.xml`) is used in the next step.

```{bash}
cp dev/interactivetool_chromatogramsvis.xml ../galaxy/tools/interactive/
```

## 6. Add the tool to the Galaxy tool panel

Galaxy only shows the tools that are listed in its tool configuration file.
Move to the Galaxy directory and edit `config/tool_conf.xml`:

```{bash}
cd ../galaxy
sed -i -e 's/askomics/chromatogramsvis/' config/tool_conf.xml
```

Galaxy's default configuration already contains an *Interactive tools* section
that lists a sample tool called `askomics`, i.e. a line similar to:

```xml
<tool file="interactive/interactivetool_askomics.xml" />
```

The `sed` command replaces the word `askomics` with `chromatogramsvis`, so the
line becomes:

```xml
<tool file="interactive/interactivetool_chromatogramsvis.xml" />
```

which points to the file copied in the previous step.

> Alternatively to overwriting the askomics IE, you can add an own entry for
> `interactivetool_chromatogramsvis.xml`

You can verify the change with:

```{bash}
grep -n chromatogramsvis config/tool_conf.xml
```

## 7. Start Galaxy

The first start takes a long time: Galaxy creates a Python virtual environment,
installs its dependencies, builds the web client and initialises its database.
Following starts take under a minute.

```{bash}
./run.sh
```

When the log shows a message similar to
`Starting server in PID ... serving on http://127.0.0.1:8080`, Galaxy is
ready.

Open <http://localhost:8080> in your web browser.

# Using the tool in Galaxy

1. Open <http://localhost:8080>.
2. Register a new account (**User > Register**) or log in. The first user is
   not automatically an administrator; this is not needed to run the tool.
3. If the tool requires input data, upload it using the **Upload** button
   (top of the left panel) and wait until the dataset turns green.
4. In the left tool panel, find the **Interactive tools** section and click on
   **ChromatogramsVis**.
5. Select the inputs/parameters, if any, and click **Run Tool**.
6. Galaxy creates a job and starts a Docker container from the image you
   built. Wait until the job is running.
7. Open **User > Active Interactive Tools**. When the *ChromatogramsVis*
   entry shows a link in the **Name** column, click it: the Shiny application
   opens in a new browser tab.
8. When you have finished, return to **Active Interactive Tools**, tick the
   tool and click **Stop**, or delete the job from the history. Galaxy then
   stops and removes the container, freeing CPU and memory.

# Updating the tool after changing the code

Docker images are not rebuilt automatically. Whenever the R code or the
`Dockerfile` changes:

```{bash}
## 1. Stop any running instance of the tool from Galaxy

## 2. Rebuild the image (from the ChromatogramsVis directory)
TOOL_VERSION=`grep Version DESCRIPTION | cut -d: -f 2 | tr -d " "`
CONTAINER_VERSION=`grep '<token name="@CONTAINER_VERSION@">' dev/interactivetool_chromatogramsvis.xml | cut -d'>' -f2 | cut -d'<' -f1`
docker build . -t chromatogramsvis:${CONTAINER_VERSION}_${TOOL_VERSION} -f Dockerfile

## 3. Start a new tool run from Galaxy
```

If you changed the XML file, copy it again into Galaxy:

```{bash}
cp dev/interactivetool_chromatogramsvis.xml ../galaxy/tools/interactive/
```

Galaxy reloads tool XML files automatically in most development setups; if the
changes do not show up, restart Galaxy (`Ctrl+C` and `./run.sh`).

If you change the **package version** in `DESCRIPTION`, make sure the image
tag referenced in the XML file still matches the image you built.

# Troubleshooting

**The tool does not appear in the Galaxy panel**: check that (i) the XML file
is in `galaxy/tools/interactive/`, (ii) the `tool_conf.xml` line points to the
right file name, (iii) Galaxy was restarted after editing `tool_conf.xml`.
Also make sure that you edited the file Galaxy is actually using
(`tool_config_file` in `galaxy.yml`).

**The Shiny app opens but shows an error**: inspect the container logs while
the tool is running or open the bash:

```{bash}
docker ps
docker logs <container_id>
docker exec -it <container_id> bash
```

# Session information

```{r session-info, eval = TRUE}
sessionInfo()
```
